Measuring contrast processing in the visual system using SSVEP

### 2024-11-27 responseToReviews.md

# Comments to the Author
## General Comments

Wade and Baker have written a very approachable and well-motivated review of the use of VEP measurements of visual contrast. The authors rightly focus on the importance of measuring contrast responses under a wide range of conditions, highlighting the ways in which human evoked responses can be made in informative in diverse contexts. The review should thus be of broad interest.

There are points where the review is overly broad and others that could use more detail. These are described below.

A central part of the review is the role of computational models and normalization models, in particular. The use of the normalization framework for describing the VEP contrast response function (CRF) has been fruitful in terms of mechanistic interpretation and linking to other measurements, such as single-unit physiology and fMRI.  This modeling approach assumes that the CRF reflects the activity of a more or less unitary population, and that the population response resembles that of the underlying individual neurons that make up the population.

There are two edge-cases in the literature that suggest a more nuanced approach will be needed. It would be useful to be clearer on the assumptions of the modeling approach and what the significance is for interpretation of cases where the assumptions don’t hold.

Two-limbed contrast response functions. Campbell and Maffei (J. Physiol,1970) in the earliest measurements of the VEP contrast response function reported that it sometimes had two “limbs” – one at low contrast and a second one with a steeper slope at higher contrasts.  Two-limbed contrast response functions have also been found by Nakayama and Mackeben (Vision Res,1982) in macaque with electrodes on the dura, by Apkarian and Tyler (Vision Res, 1985),  Norcia et al., (Vision Res,1989) and Souza et al., (IOVS, 2007), to name a few.  As noted, the sigmoidal model assumes that the population response is unitary, but the two-limbed functions suggest that it is a mixture of populations. The two limbs may not be apparent when the function is sampled with only a few points over a wide range of contrast.

Harmonics with different CRFs.  Bobak et al., (IOVS, 1986), found that the first harmonic (1f) and second harmonic (2f) of the pattern onset/offset SSVEP had different thresholds but similar slopes once above threshold. Norcia et al., (TVST, 2024) found different thresholds and slope for 1f and 2f responses driven by contrast modulated dynamic noise. This again suggests the presence of multiple underlying populations.  The behavior also raises the question of whether to pool harmonics when estimating the response function. This was alluded to in the simulation, but could be expanded on.

The lack of dynamics in the normalization model, as presented should be discussed, the brief mention of Tsai et al. aside. The authors may wish to cite another framework that explicitly models dynamics (e.g. amplitude and phase) in the context of the contrast response function (Zemon and Gordon, Vision Res, 2006) .

### Response
> We thank the reviewer for these positive comments. In general this was intended to be an overview in the sense of a 'primer' for SSVEP rather than a detailed historical account of the field - hence the inclusion of code examples. However, the reviewer raises some important general issues and we now address them: Specifically, we include references to the idea that SSVEP can engage more than one type of neuron, that the population responses is not trivially derived from what is known about the properties of individual neurons and that fitting with a sigmoidal response function may mask some complexities in the CRF that are more readily seen with denser sampling. We think the Bobak reference is IOVS 1984 and we have also included the Kaestner TVST 2024 reference (thank you for drawing our attention to both of these).

### Specific comments

Adaptation effects.  Adapted contrast response functions have been measured by Nelson et al., (EEG Journal, 1984) by comparing CRFs measured by sweeping contrast from low to high vs high to low. They found that responses at low contrast were smaller in the “down” sweeps and attributed it to differences in prior adaptation.  Other studies also measured adaptation effects on the CRF Heinrich and Bach, (IOVS, 2001), Ross and Speed (1991, Proc Roy Soc). And the hard-to-find Bach, Greenlee and Buhler (Clin Vis Sci, 1996) paper that modeled the effect within a contrast gain control framework.
Gain control development.  Morrone and Burr (Nature, 1986) used frequency tagging to measure orientation dependence of masking in infants and adults, as did Candy et al. The difference in the two studies is that in adults Morrone and Burr found response gain for cross-oriented maskers but Candy et al found contrast gain.   Skoczenski and Norcia used a high contrast 30 Hz dynamic 2D noise masker and variable contrast reversing gratings. They found masking consistent with a contrast gain effect.

>

Page 5 bottom. The reciprocal effect of test and masker was first shown by Regan and Regan  (1988; cited)  for flicker modulation depth and then for gratings by Brown et al., (IOVS,  1999 for dichoptic masking) and  then  by candy et al., (J. Neurosci, 2001) along with Busse et al. (Neuron, 2009) as cited. It’s good to have cited Regan and Regan paper that introduced the notion of fingerprints in the spectrum being tied to the nature of the non-linearity. Maybe expand their “fingerprint” approacg a little more.

Figure 4. Oversaturation occurs because the denominator overtakes the numerator at high contrast as it has two terms in it, not the single term in the numerator. Would be worth showing total power not saturating. Is this a general effect?

Equation 3 is a special case where the neurons in the numerator are so selective that they don’t “see” the masker except by being influenced indirectly by the gain pool.  Putting the test and mask in the numerator and summing them before exponentiating yields robust IM. Weights can be used to reflect the neurons preference for the test vs the mask.  Does equation 3 produce IM?

Page 7 Mckeefry et al., (Vis Neuro, 1996) is a good reference for linking harmonics of on/off to transient sustained responses.

Page 8 figure 5. The extrapolation to zero amplitude method should be attributed to Cambell and Maffei (J Physiol) 1970). Campbell and Kulikowski (J Physiol, 1972) made a detailed comparison to psychophysics.

Figure 5 is an oversimplification. As it doesn’t take into account how the additive noise mixes with the internal VEP response.  This was modeled in Norcia et al., (Vision Res, 1989). The additive noise adds a bias to the measurement, especially if the slope of the response function is shallow. The authors should try to connect the linear regression approach for threshold measurement to the sigmoidal non-linearity, as they are different models of the CRF. In practice the difference at the low contrast end is probably too small to matter and the saturation region is excluded.

Page 9. The sweep method was first developed by David Regan (Vision Res, 1973) for use in objective refraction.

Page 9 “This approach” seems to refer to the sweep VEP, not the method of measuring at multiple contrast levels over a series of trials which was used in the Atkinson/Braddick papers. The sweep method refers to the approach of incrementing or decrementing the stimulus value within a single trial and then averaging the sweeps to get the response function.  Regan argued that this was a better way of estimating response functions because amplitude variations over a session are distributed over the whole response function shifting it up or down instead of distorting its shape (Regan (Vison Res, 1975, color coding).

Page 9 …chromatic and chromatic contrast as well as stereoscopic depth perception develops earlier than had been supposed previously based on behavioural readouts (Dobson et al., 1978; Norcia & Tyler, 1985b) with both chromatic and achromatic contrast detection reaching near-adult levels by around six months. Needs references for chromatic sensitivity and disparity sensitivity.

The DeVries Khoe and Spekreijse comparison between VEP and psychophysics was done above age 4 years and for acuity not for contrast sensitivity. The large discrepancies between VEP and FPL are in infancy for contrast sensitivity, as well as for acuity.  In the realm of contrast sensitivity, the differences are very large, up to a factor of 50.  Compare Norcia et al., 1990 Vision Res measurement to any behavioral data. Suggest making this comparison more quantitative.

The authors may wish to cite Norcia et al., (NeuroReport, 2000) for an early measurement of suppression in amblyopia using measurements of the contrast response function in each eye in the presence of a high contrast dichoptic masker in the other.  This study showed that amblyopic observers retain suppressive binocular interactions despite severely reduced stereopsis. The dominant eye exerted a greater suppressive effect on the non-dominant eye, suggesting that it maintains a relatively greater number, or more effective, inhibitory contacts with cortical cells.  Also may wish to cite recent work by Hou and colleagues using a similar approach, including application of normalization models.

Schade (JOSA, 1956) was the first to measure the contrast sensitivity function of the human visual system as noted by Campbell and Robson (J. Physiol, 1968). Campbell and Green (1965, cited) credit Schade for how to make sinewave gratings on an oscilloscope but not for having made the first measurement.

Villadaite, Norcia et al., (cited) did not find spatial frequency specific effects at 8 cpd as in Pei et al,. (cited). So it’s not really a replication of that study, but more of an extension to a measurement of CRF instead of SF tuning.

Page 12, better to state whether attention effects were consistent with response gain or contrast gain to make the result more concrete for the reader.

Discussion of time varying response paragraph on page 12 is not that relevant to contrast response function measurement, nor is the BCI section.  These applications of SSVEP and could be deleted or the relationship to contrast response function made clearer.

Page, 14. It’s not clear from the text how one would study feedback with the SSVEP, please elaborate.

In the last section, it would be worth mentioning that in order to “identify” a non-linear system, one needs multiple inputs. If the system is 2nd order, a minimum of two inputs is needed. If the system non-linearity is higher order than that, then more simultaneous inputs are needed (see Boyd, Tang and Chua 1983, IEEE Trans Circuits and Systems  and  Chua and Liao, 1991, Int. J. Circuit Theory and Applications).

.

## Reviewer: 2

# Comments to the Author
This manuscript is intended to be a review of the field of steady-state visual evoked potentials (ssVEPs) over the past 70 years.  The authors state in the introductory paragraph “Here we will describe how an EEG method known as the steady state visually evoked potential (SSVEP) technique has contributed to our understanding of human contrast processing in health, disease and throughout development.” The material included and the references cited, however, are quite limited and omit seminal work in this area.

1. All abbreviations should be defined upon first use.
2. The following statement is unclear and inaccurate: “On-off flicker can drive independent populations of on- and off-cells in the retina once per cycle and can therefore produce a response at the fundamental flicker frequency, known as 1F, and its integer harmonics: 2F, 3F, 4F and so on. Counterphase flicker contains two transients per cycle and therefore does not produce a response at 1F, only at its even harmonics: 2F, 4F, 6F and so on.” The term “on-off flicker” is typically used for simple luminance flicker without spatial pattern contrast. In the context of pattern stimulation, “on-off” can refer to “appearance-disappearance” of a spatial pattern with its contrast modulated in time. When the pattern appears and disappears into a uniform field of equal space-average luminance, the light and dark elements are modulated in counterphase. Any linear contribution to the response elicited by each set of elements (e.g., bars or checks) is expected to be canceled when summed at the electrode site on the scalp. Thus, the response at 1F to this stimulus is in fact a nonlinear response. This type of stimulation does not separate contributions from “independent populations of on- and off-cells.” There are other ssVEP techniques developed that do tap ON and OFF pathways selectively, which are not cited here. Contrast-reversal stimuli which involve symmetrical modulation of light and dark elements cancel odd harmonic frequency components in the response, however, both kinds of stimulus techniques mentioned here elicit only nonlinear responses.
3. The statement “SSVEPs have proven to be an excellent measure of early chromatic processing as well” again fails to cite seminal work in this area.
4. The statement “This normalization, achieved through a computation called ‘contrast gain control’” again neglects to cite the original work that defined “contrast gain control” and the authors ignored a nonlinear biophysical model proposed to explain this phenomenon in ssVEP contrast response functions (which was also applied in developmental work on infants and children).
5. The description of Figure 3 and the nonlinear operations performed are unclear. They state in the text that Equation 2 (“hyperbolic ratio function”) is used as a nonlinear transducer, but the figure legend states that a squaring operator was used for frequency doubling. A squaring operation would only produce a sinusoid of double the frequency of the input. The waveform shown in the left panel appears to be a full-wave rectified signal, which would contain theoretically an infinite number of higher harmonics. The text also states that “This results from the distortion of the input sine waves at high contrast due to a combination of the full-wave rectification and saturating non-linearity.” When was full-wave rectification performed?
6. The authors use the term “frequency tagging” which has been used in the literature to refer to stimulating different regions of the stimulus field or fellow eyes with different sinusoidal frequencies and measuring the intermodulation (sum and difference) frequency responses that result from nonlinear interactions. They do not, however, cite the foundational work in this field.
7. There is a large body of work that used ssVEPs to investigate contrast processing developmentally (infants through adulthood) and in various disease states, which is mostly missed in this review article.
8. Figures should be labeled.
Thus, the stated intention of this manuscript (“Here we describe how SSVEPs have been used to study visual contrast over the past 70 years”) fails to be met. Only select articles from a few sources are covered, and therefore, the richness and value of this technique is lost.

### 2024-11-27 review.qmd
Editing review

Although its mathematical definition is straightforward, the computations that underlie contrast processing in the brain have been the subject of intense research for many decades. The neural code for contrast, even in the earliest parts of visual cortex, is not simply a linear transform of the contrast at the retina - instead, contrast signals undergo a cascade of nonlinear processing stages that, broadly, attempt to normalise the output relative to the spatiotemporal environment. This normalization, achieved through a computation called 'contrast gain control' [@Foley1994; @Heeger1992; @Carandini2011] maximises the sensitivity of the visual system by making optimal use of neuronal bandwidth. As an example, a grating placed at the centre of a low-contrast background typically appears more intense than the same grating when superimposed on a high contrast background (see @fig-centresurround; note that the code used to produce all figures in this review is available in python and R at: https://github.com/wadelab/contrastReviewPaper).

@@

>>>>

Although its mathematical definition is straightforward, the computations that underlie contrast processing in the brain have been the subject of intense research for many decades. The neural code for contrast, even in the earliest parts of visual cortex, is not simply a linear transform of the contrast at the retina - instead, contrast signals undergo a cascade of nonlinear processing stages that, broadly, attempt to normalise the output relative to the spatiotemporal environment. This normalization, achieved through a computation called 'contrast gain control' [@Foley1994; @Heeger1992; @Carandini2011] maximises the sensitivity of the visual system by making optimal use of neuronal bandwidth. As an example, a grating placed at the centre of a low-contrast background typically appears more intense than the same grating when superimposed on a high contrast background (see @fig-centresurround; note that the code used to produce all figures in this review is available in python and R at: https://github.com/wadelab/contrastReviewPaperVNS).

@@

An additional complexity is introduced by the fact that the EEG is an average population response and that the visual system contains many different types of neurons. As a stimulus changes in, for example, contrast it may selectively activate qualitatively different neuronal populations. This type
